Variation in distance between consecutive animal locations

### 2021-04-13 distributions_paper/rudolph_distributions.Rmd
updates to the introduction

# Introduction
   Movement is a fundamental ecological process, critical to an animal's biology and it's interactions with the environment. The study of animal movement tends to focus on four basic mechanistic questions: the why, when, where, and how an animal moves, and these questions can be answered at the scale of individuals or populations. A general framework for movement ecology is needed to link movement patterns and processes. Such a framework would start with a focus on movement itself and the individual's internal state, navigation capacity, motion capacity, and responses to external factors [@nathan2008movement]. The overall goal of studying movement is to find the linkages between the causes and mechanisms of animal movement, that explain spatiotemporal patterns that then feed back into various ecological and evolutionary processes. Some of these approaches involve separating different movement patterns into states, with each state having an associated set of parameters that determine that specific movement pattern [@langrock2012flexible]. These approaches segment movement patterns, in order to associate specific types of movement to underlying behaviors or fine-scale environmental influences (Getz citations, Morales, more focus on the random walk models, Patterson2008). Several of these approaches focus on the why and underlying processes that drive animal movement, whether that is an animal's internal state or responses to their environment.

   Random walks have been used broadly in biology to characterize animal movement, and increasing levels of complexity can be incorporated by using multiple random walks, continuous time or different levels of correlation between turning angles [@codling2008random; @morales2004extracting]. More recent developments have focused on implementing higher order Markov processes through state-space models, which provide a probabilistic framework that allows for prediction of future states of a system based from previous states, by coupling an observation model with an underlying mechanistic model. When extending this framework to movement ecology, we understand that the process model provides the mechanism for movement across space and time, and the observation model is related to the sampling process. In movement data, a state-space process allows us to couple a statistical model for the observation method, accounting for errors in sampling or detection, with a separate model for the movement dynamics which are determined by the effects of behavior or responses to the environment [@patterson2008state]. In general, state-space models are used to identify different states, or behavioral states, that determine different movement patterns associated to step length and turning angle. And some of the main questions asked with this approach are related to identifying these different states, or behavioral state-switching, from time series data on location tracking.

   Although I have mostly described approaches used to understand why movement happens, animal movement also has consequences, which might influence species interactions, ecological function, and vital demographic processes (such as survival, reproduction, or dispersal of other species) and thus result in population or regional level impacts. A clear example of the consequences of animal movement on other species involves animal-mediated seed dispersal or pollination, impacting a plant's spatial population dynamics.The specific case of seed dispersal is interesting because of the different levels and interactions happening. At the individual level, we focus on an individual bird consuming fruit and dispersing the seeds of the tree. How this particular individual moves across the landscape directly affects the dispersal for that specific tree and other trees from which it consumes fruit. At the same time, that individual has intra-individual variation in movement patterns associated to specific events in that individual's lifetime, such as breeding and nesting, or even smaller scale variation related to different states, such as foraging or resting. Interspecific interactions come at play as multiple species find themselves together feeding from the same trees, and thus competing for these resources, while at the same time intraspecific interactions due to social behavior broadly determine general movement patterns for animals restricted within family home ranges. From the plant's perspective, the variation in movement patterns across all dispersers will determine eventual plant population spatial spread.

  Specifically in the context of seed dispersal ecology, the general focus has been on understanding how the typical movement patterns, built around average measures at population levels, can influence plant population spread and seed dispersal [sources]. Animal movement can be modeled with a simple diffusion process or a random walk [@jones2017closing; @levey2005effects], with the simplest approach assuming model parameters are the same across individuals, something known as 'complete pooling' [@langrock2012flexible]. This focus on generality has driven to a lack of frameworks for understanding how variation in animal movement is maintained, and the consequences of this variation across populations and communities. From the animal movement perspective, some alternatives have been considered to incorporate individual variation with 'no pooling' or 'partial pooling' approaches, where each or some individuals have their own set of parameters [@patterson2009classifying, @jonsen2006robust], however a high number of parameters quickly limits applications to high levels of individual variation. Recent calls for incorporating individual variation in movement and their consequences to seed dispersal or even pathogen spread [@snell2019consequences, @shaw2020causes] have pushed for the need to develop theory that explores different types of variation in movement patterns to understand its consequences across ecological organizational levels. In particular, when considering species interactions, incorporating individual variation becomes essential to understand how positive or negative feedback loops play into the maintenance of variation in movement as individuals respond to external factors in their environment [@shaw2020causes].

  We focus here on developing a process to analyze animal movement sequences based on movement lengths, or step lengths, and turning angle probability distributions. Following the methods of Kareiva and Shigesada [-@kareiva1983analyzing], we assume the movement is a correlated random walk and derive expected square displacements associated to a number of consecutive moves under various probability distribution models.A simple correlated random walk, which involves a directional correlation between successive step movement [@codling2008random]. Unlike other studies which focus on deciphering the underlying behaviors from animal movement data, our focus was to uncover potential differences between individuals and how these differences can be obscured when modeling population-level animal movement. We know these simple correlated random walks, or simple movement models might not be enough to describe net displacements under more complex processes, for which developments in higher order markov processes show to be promising. However, our goal in this study was to use a simpler model and focus on potential differences when analyzing complete pooling versus no pooling data. In the context of animal-mediated seed dispersal, understanding these indivdiual level differences can help us describe and characterize the implications that different animals may have over spatial distributions of plant populations, and thus identify potential dispersers with higher or more importance, particularly when associated to long-distance seed dispersal events.

@@

We focused our study of individual variation on the many-banded araçari (*Pterglossus pluricinctus*), a small toucan in Yasuní National Park, Ecuador. We used previously collected data from studies using radiotelemetry to estimate home ranges and evaluate the potential seed dispersal distances for the *Virola flexuosa* tree [@holbrook2011home, holbrook2007usin]. Over a period of four years, from 2001 to 2005, *P. pluricinctus* individuals were captured and radio-transmitters were attached at the base of the central tail feathers [@holbrook2011home]. Toucan locations were estimated by triangulation using receivers and hand-held antennas, with tracking periods lasting between 4-5 hours per day, locating individual birds every 15 minutes. Further details on field methods can be found in Holbrook [-@holbrook2011home]. It is important to note that these methods did not track all birds simultaneously, and that although individual toucan locations were attempted every 15 minutes, this was not always possible, thus some successive locations have time categories associated to multiples of 15. From the relocation data, we calculated animal trajectories by considering discrete steps between successive relocations of each animal [@turchin1998quantitative], quantifying their turning angle after each movement, estimating the Euclidean distance between successive relocations, and estimating the velocity for each segment as the displacement distance divided by the length of the time interval. We used the package 'adehabitatLT'[@adehabitatLT] in R [@Rlang] to calculate the descriptive parameters for each trajectory, such as turning angles and net squared displacement.

@@

@@

```{r setup, include=FALSE}
# knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
knitr::opts_chunk$set(message = FALSE, warning = FALSE, 
                      #eval=FALSE,
                      echo = FALSE)

@@

>>>>

# Introduction

Movement is a fundamental ecological process, critical to an animal's biology and it's interactions with the environment. The study of animal movement tends to focus on four basic mechanistic questions: the why, when, where, and how an animal moves, and these questions can be answered at the scale of individuals or populations. A general framework for movement ecology is needed to link movement patterns and processes. Such a framework would start with a focus on movement itself and the individual's internal state, navigation capacity, motion capacity, and responses to external factors [@nathan2008movement]. The overall goal of studying movement is to find the linkages between the causes and mechanisms of animal movement, that explain spatiotemporal patterns that then feed back into various ecological and evolutionary processes. However, we can also focus on the effects and consequences that different movement patterns can have over species interactions and other ecological processes.

Random walks have been used broadly in biology to model movement of animals and microorganisms, starting with the simplest models of movement using uncorrelated and unbiased random walks [@codling2008random]. These movement models show no preference in direction and movement at each step is completely independent of the direction taken in previous movements. Extensions of simple random walk processes are regularly used to accommodate for more complex behaviors or to include drift. Some of these extensions involve the use of correlated random walks, which incorporate a correlation between directions in successive movements, biased random walks, which incorporate a directional bias, or even building movement models as mixtures of random walks to account for more complex behaviors [@morales2004extracting]. Modeling approaches using random walks generally assume a homogeneous environment and behaviors that don't drastically change the movement pattern [@kareiva1983analyzing]. A simple random walk approximates displacements by connecting a series of straight lines, and can summarize animal movement based on parameters for movement length and turning angle, or the use of net squared displacement. Connecting theoretical movement models based on correlated random walks to data had often relied on the use of these parameters or displacement measures [@turchin1998quantitative], but more current and efficient approaches have used likelihood-based techniques to estimate parameters by fitting a discrete approximation of the model directly to the location data [@jonsen2003meta].

   In the context of species interactions and the consequences of animal movement to dispersal, the general focus has been on understanding how the typical movement patterns, built around average measures at population levels, can influence plant population spread [@levey2005effects], seed dispersal [@jones2017closing, @holbrook2007using], pollen dispersal [@garcia2007contemporary], and even pathogens [@shaw2017vector]. Animal movement can be modeled with a simple diffusion process or a random walk [@jones2017closing; @levey2005effects], with the simplest approach assuming model parameters are the same across individuals, something known as 'complete pooling' [@langrock2012flexible]. This focus on generality has driven to a lack of frameworks for understanding how variation in animal movement is maintained, and the consequences of this variation across populations and communities. From the animal movement perspective, some alternatives have been considered to incorporate individual variation with 'no pooling' or 'partial pooling' approaches, where each or some individuals have their own set of parameters [@patterson2009classifying, @jonsen2006robust], however a high number of parameters quickly limits applications to high levels of individual variation. The decision to pool data or not may also come from the modeling framework chosen, as frameworks that allow for random effects may have a better structure to incorporate this variation than simpler frameworks. Recent calls for incorporating individual variation in movement and their consequences to seed dispersal or even pathogen spread [@snell2019consequences, @shaw2020causes] have pushed for the need to develop theory that explores different types of variation in movement patterns to understand its consequences across ecological organizational levels. In particular, when considering species interactions, incorporating individual variation becomes essential to understand how positive or negative feedback loops play into the maintenance of variation in movement as individuals respond to external factors in their environment.

  We focus here on developing a process to analyze animal movement sequences based on movement lengths, or step lengths, and turning angle probability distributions. We assume the movement is a correlated random walk and derive expected square displacements associated to a number of consecutive moves under various probability distribution models. Unlike other studies which focus on deciphering the underlying behaviors from animal movement data, our focus was to uncover potential differences between individuals and how these differences can be obscured when modeling population-level animal movement. We know these simple correlated random walks, or simple movement models might not be enough to describe net displacements under more complex processes, for which developments in higher order markov processes show to be promising. However, our goal in this study was to use a simpler model and focus on potential differences that arise when comparing data from multiple individuals under frameworks that completely pool data in contrast to those that don't.

@@

## Animal movement model

## Description of the models considered

## Specific case with telemetry data

We focused our study of individual variation on the many-banded araçari (*Pterglossus pluricinctus*), a small toucan in Yasuní National Park, Ecuador. We used previously collected data from studies using radiotelemetry to estimate home ranges and evaluate the potential seed dispersal distances for the *Virola flexuosa* tree [@holbrook2011home, holbrook2007usin]. Over a period of four years, from 2001 to 2005, *P. pluricinctus* individuals were captured and radio-transmitters were attached at the base of the central tail feathers [@holbrook2011home]. Toucan locations were estimated by triangulation using receivers and hand-held antennas, with tracking periods lasting between 4-5 hours per day, locating individual birds every 15 minutes. Further details on field methods can be found in Holbrook [-@holbrook2011home]. It is important to note that these methods did not track all birds simultaneously, and that although individual toucan locations were attempted every 15 minutes, this was not always possible, thus some successive locations have time categories associated to multiples of 15. Due to the nature of the data, and the irregular time lags between two consecutive locations, previous work calculated movement rate for each individual in order to take advantage of all the data despite variable time intervals [see Table 1 in @holbrook2011home]. Movement rates were calculated as the displacement in meters over the total length of the tracking period in minutes, where the tracking period was defined as the total active tracking time for a bird over the season when the data was collected. In other words, the movement rate could be defined as the average velocity of the animal over the entire tracking period, and thus there is one movement rate value associated to each individual bird. 
In our case, we wanted to extend the use of the telemetry data and use all locations in a modeling framework. From the relocation data, we calculated animal trajectories by considering discrete steps between successive relocations of each animal [@turchin1998quantitative], quantifying their turning angle after each movement, estimating the Euclidean distance between successive relocations, and estimating the velocity for each segment as the displacement distance divided by the length of the time interval. We used the package 'adehabitatLT'[@adehabitatLT] in R [@Rlang] to calculate other descriptive parameters for each trajectory, such as turning angles and net squared displacement.

@@

What do we consider a trajectory? The path followed by consecutive relocations during a tracking session. The tracking session for this data is the 4-5 hour period each day that they follow the bird. Then, the number of relocations for each trajectory varies from one to 26, with a mean of 4 relocations. This is a limitation of the data, since it's not easy to characterize a full trajectory of movement with such a small number of relocations in a trajectory.

Dealing with the issue of autocorrelation in this case: what does that mean? Or we just assume each step is independent from the others. Think I saw this in the Kareiva Shigesada paper. 
```{r eval=FALSE}
# For this to run I need to move the load of data and such higher up
aracari_df %>%
  group_by(id, burst) %>% 
  tally() -> relocs

# Having a minimum number of relocations for "time series analysis" or something, fitting.
relocs %>% 
  filter(n >= 5) %>% 
  distinct(id)

relocs %>% 
  ggplot(., aes(x = n, fill = id)) +
  geom_histogram()
summary(relocs$n)
```

```{r eval=FALSE}
aracari_df %>% 
  filter(R2n >0) %>% 
  ggplot(., aes(x = R2n)) +
  geom_histogram() +
  facet_wrap(~id)
```

@@

```{r setup, include=FALSE}
# knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
knitr::opts_chunk$set(message = FALSE, warning = FALSE, 
                      eval=FALSE,
                      echo = FALSE)

@@

```{r eval=FALSE}
# Table replicating results from Holbrook 2011, movement rates
# The values are close, but not the same, so I don't know how Kimberly estimated her movement rates

aracari_df %>%
  drop_na(dist) %>% 
  mutate(mpm = dist/(dt/60)) %>% 
  group_by(id) %>% 
  summarise(movrate = mean(mpm)) %>% 
  ungroup() %>% 
  summarise(mean = mean(movrate),
            sd = sd(movrate),
            median = median(movrate),
            range = range(movrate))

aracari_df %>%
  drop_na(dist) %>% 
  group_by(id) %>% 
  summarise(movrate = sum(dist)/sum(dt/60)) %>% 
  ungroup() %>% 
  summarise(mean = mean(movrate),
            sd = sd(movrate),
            median = median(movrate),
            range = range(movrate))
